---
title: "vis_LEC_tumor_skin"
editor: visual
---

# Visualize expression across different conditions (LECs in tumor and skin)

The experiment consist of 3 main factors of interest:

-   sample origin: tumor, skin, lymph node

-   main cell type class: leucocytes, LECs

-   tumor type: YUMM, YUMMER

As a first check we are interested in the expression of specific marker genes/known marker and how they vary across combinations of these factors.

## Preamble

```{r}
#| label: load-libs
#| echo: true
#| output: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(patchwork)
library(Seurat)
library(stringr)
library(ggthemes)
library(forcats)
library(gridExtra)

library(ggrepel)
```

## Data objects

From the FGCZ we have different data objects:

1.  Tumor LECs integrated (only tumor LECs, but have been integrated with skin samples) (`data/scData_LEC_tumor.rds`)

2.  Tumor + skin LECs integrated (`data/scData_LEC_tumor_skin.rds`)

3.  All skin + tumor cells

4.  All lymphnode cells

Here we work with **object 3 - Integrated tumor and skin LECs**.

```{r}
#| label: data

#### ----- change to local path to scData.rds object here ----- #####
# e.g. seurat <- readRDS("/home/project/data_folder/scData.rds)")
seurat<- readRDS(file.path("..", "..","..","data", "scData_all_tumor_skin.rds"))
#### --------------------------------------------------------- #####


# correct condition assignment!!
seurat$cond <- seurat[[]] |> 
  mutate(
    cond = case_when(
       str_detect(Sample, "YUMM[0-9]") ~ "YUMM",
       str_detect(Sample, "YUMMER") ~ "YUMMER",
       str_detect(Sample, "Skin") ~ "skin"
    )
  ) |> select(cond)

# check assignment
table(seurat$Sample, seurat$cond)
table(seurat$cond)

seurat$cell_types <- seurat$ident |> 
  as.factor() |> 
  forcats::fct_collapse("monocytes/macrophages" = c("0","1","2","5","6","18","22"),
                        "DCs" = c("4", "12", "19"),
                        "CD4 T cells" = c("10"),
                        "CD8 T cells" = c("7", "16"),
                        "other T cells" = c("3", "11", "15", "23", "24"),
                        "endothelial cells/fibroblasts" = c("8", "9", "13"),
                        "mast cells" = c("14"),
                        "B cells" = c("21"),
                        "other" = c("17", "20"))


DefaultAssay(seurat) <- "SCT"

cluster_marker <- c("Cd3e", "Cd8a", "Cd4" ,"Tcf7", "Lef1", "Pdcd1", "Sell", "Tigit", "Cxcr4", "Ccr7", "Foxp3", "Ctla4", "Trdc", "Il12b", "Itgae", "Xcr1", "Cd207", "Notch2", "Itgax", "Itgam", "Il1a", "Cd80", "Cd19", "Prox1", "Pecam1", "Csf2rb", "Csf3r", "Il1r2", "Acta2", "Pdgfrb", "Ctsk", "Cd74", "H2-Ab1", "H2-Aa", "Krt15", "Krt5", "Cd79a", "Ccl3", "Ccl4", "Cd63")

cond_marker <- c("Cd3e", "Cd8a", "Cd4", "Tcf7", "Lef1", "Pdcd1", "Foxp3", "Cxcr4")
```

## Overview

#### Mean features/molecules per condition

```{r}
#| label: mean feature/mol counts

# mean molecules/group
mean_mol <- seurat[[]] |> group_by(cond) |>  summarise(mean_mol = mean(nCount_RNA))
mean_mol

# mean unique genes/group
mean_feature <- seurat[[]] |> group_by(cond) |>  summarise(mean_gene = mean(nFeature_RNA))
mean_feature

```

### Distribution of conditions per cluster

```{r}
#| lable: cond-dist

cond_dat <- seurat[[]] |> group_by(cond, cell_types) |>  summarise(n_cells = n())

ggplot(cond_dat, aes(fill=cond, y=n_cells, x=cell_types)) + 
  geom_bar(position="stack", stat="identity") + 
  scale_fill_tableau() + 
  theme_bw()

ggplot(cond_dat, aes(fill=cond, y=n_cells, x=cell_types)) + 
  geom_bar(position="fill", stat="identity") + 
  scale_fill_tableau() + 
  theme_bw()

ggplot(cond_dat, aes(fill=cell_types, y=n_cells, x=cond)) + 
  geom_bar(position="fill", stat="identity") + 
  scale_fill_tableau() + 
  theme_bw()

```

### Distribution of cells within umap

```{r fig.width=9, fig.height=4}
#| label: dimplot cond

DimPlot(seurat, group.by = "ident", 
        split.by = "cond", 
        reduction = "umap", 
        label = T) + 
  ggtitle("") + 
  theme(legend.position = "none")


#Almut please put number of clusters inside the plot
```

### Distribution of broad cell types

```{r, fig.width=12, fig.height=4}
#| label: broad ct umap

DimPlot(seurat, group.by = "cell_types", 
        split.by = "cond", 
        reduction = "umap", 
        label = T) + 
  ggtitle("") 

```

## Cluster marker gene expression

+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **Cluster**                   | Genes                               | **Cluster**                | Genes                                | **Cluster**           | Genes                           |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **0 - Monocytes/\             | Cxcr4, Notch2, Itgax, Csf2rb, Il1r2 | **9 - LECs**               | Prox1, Pecam1                        | **18 - Macrophages**  | Cxcr4, Il1a, CD80, Csf3r, Il1r2 |
| Macrophages**                 |                                     |                            |                                      |                       |                                 |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **1 - Monocytes/\             | Cxcr4, Notch2, Itgam                | **10 - Tregs**             | Cd3e, Cd4, Sell, Tigit, FoxP3, Ctla4 | **19 - DCs**          | Cd74, H2-Ab1, H2-Aa             |
| Macrophages**                 |                                     |                            |                                      |                       |                                 |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **2 - Granulocytes**          | Notch2, Itgax, CD80, Csf2rb, Il1r2  | **11 - TGD**               | Cd3e, Tcf7, Trdc                     | **20 -Keratinocytes** | Krt15, Krt5                     |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **3 - TGD**                   | Tcf7, Lef1, Sell, Tigit, Trdc       | **12 - DC**                | CCR7, Il12b, CD80, Csf2rb            | **21 - B cells**      | CD19, Cd79a                     |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **4 - cDC2**                  | Notch2, Itgam                       | **13 - Fibroblasts**       | acta2, Pdgfrb                        | **22 - Monocytes**    | Itgax                           |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **5 - Macrophages**           | Notch2, Itgax, Itgam, Csf2rb        | **14 - Mast cells**        | Notch2, Csf2rb, Cd63                 | **23 - TGD**          | Tcf7, Trdc                      |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **6 - Monocytes**             | Itgam                               | **15 - T cells**           | Cd3e, Tcf7, Cxcr4,                   | **24 - T cells**      | Ccl3, Ccl4                      |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **7 - exhausted CD8 T cells** | Cd3e, Cd8a, Pdcd1                   | **16 - naive CD8 T cells** | Cd3e, Cd8a, Tcf7, Ctla4              |                       |                                 |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+
| **8 - LECs**                  | Notch2, Prox1, Pecam1               | **17 - Osteoclasts**       | Ctsk                                 |                       |                                 |
+-------------------------------+-------------------------------------+----------------------------+--------------------------------------+-----------------------+---------------------------------+

### Dimplots

```{r fig.width=12, fig.height=35}
#| label: dimplots

FeaturePlot(object = seurat, features = cluster_marker, ncol = 3)

```

### Violinplots

```{r, fig.width=14, fig.height=50}
#| lable: violin marker
Idents(seurat) <- "cell_types"
VlnPlot(object = seurat, features = cluster_marker, ncol = 3, assay = "SCT", add.noise = F)
```

### Dotplots

```{r fig.width=13, fig.height=5}
#| label: dotplot

DotPlot(object = seurat, features = cluster_marker) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```

## Conditional marker expression

### Expression by condition as Dimplot

```{r, fig.width=10, fig.height=25}
#| label: dim cond

FeaturePlot(object = seurat, features = cond_marker, ncol = 3, split.by = "cond")

```

### Expression as violin plot grouped by cluster

```{r, fig.width=8, fig.height=30}
#| label: vln cond cluster
#| warning: FALSE

VlnPlot(object = seurat, 
        features = cond_marker, 
        assay = "SCT", 
        add.noise = F, 
        split.by = "cond",
        ncol = 1) + 
  plot_layout(guides = 'collect') &
  scale_fill_tableau() &
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

```

### Expression as violin plot clusters combined

```{r, fig.height=12, fig.width=8}
#| label: vln cond only
#| warning: FALSE

VlnPlot(object = seurat, 
        features = cond_marker,
        group.by = "cond",
        assay = "SCT", 
        add.noise = F,
        ncol = 2) + 
  plot_layout(guides = 'collect') &
  scale_fill_tableau() &
  theme_bw()

```

### Expression as violin plot cluster 7

```{r, fig.height=12, fig.width=8}
#| label: vln cond7 only
#| warning: FALSE
Idents(seurat) <- "ident"

# Subset your Seurat object to only include cells from cluster 7
cluster7_seurat <- subset(seurat, idents = 7)

# Create the violin plot for cluster 7
VlnPlot(object = cluster7_seurat, 
        features = cond_marker,
        group.by = "cond",
        assay = "SCT", 
        add.noise = F,
        ncol = 2) + 
  plot_layout(guides = 'collect') &
  scale_fill_tableau() &
  theme_bw()
```

### Expression as violin plot cluster 16

```{r, fig.height=12, fig.width=8}
#| label: vln cond716 only
#| warning: FALSE

# Subset your Seurat object to only include cells from cluster 16
cluster16_seurat <- subset(seurat, idents = 16)

# Create the violin plot for cluster 7
VlnPlot(object = cluster16_seurat, 
        features = cond_marker,
        group.by = "cond",
        assay = "SCT", 
        add.noise = FALSE,
        ncol = 2) + 
  plot_layout(guides = 'collect') &
  scale_fill_tableau() &
  theme_bw()
```
