---
title: "compare tumor YUMM"
format: html
editor: visual
---

## Sequencing comparison

Compare cells from different sequencing runs.

### Library

```{r}
#| label: libs
#| echo: true
#| output: false
library(dplyr)
library(batchelor)
library(scater)
library(scran)
library(SingleCellExperiment)
library(patchwork)
```

### Data

Objects combine sequencing runs within tumor or lymphnode samples and tumor type (YUMM/YUMMER)

OBS: Objects are large run separately and clear complete memory afterwards, if run locally. Uncomment relevant lines.

```{r}
#| label: data

sce <- readRDS("../../data/seq_Nov/objects/sce_tumor_yumm_combined.rds")
#sce <- readRDS("../../data/seq_Nov/objects/sce_ln_yumm_combined.rds")
#sce <- readRDS("../../data/seq_Nov/objects/sce_tumor_yummer_combined.rds")
#sce <- readRDS("../../data/seq_Nov/objects/sce_ln_yummer_combined.rds")


```

### Cell numbers

##### Prox1 or Pecam1 & Pdpn expressing cells

```{r}
#| label: celln

cell_gene_count <- function(sce, gene){
  length(which(counts(sce)[gene, ] > 2))
}

cell_gene_comb <- function(sce, gene1, gene2, gene3){
  length(which(counts(sce)[gene1, ] > 2 & counts(sce)[gene2,] > 2 | 
                 counts(sce)[gene3, ] > 2))
}

print("Number cells with >2 Prox1 in both experiments:")
cell_gene_count(sce, "Prox1")

print("Number cells with >2 Prox1 OR >2 Pdpn AND Pecam1 in both experiments:")
cell_gene_comb(sce, "Pdpn", "Pecam1", "Prox1")

print("Number cells with >2 Prox1 in old experiments:")
cell_gene_count(sce[, sce$experiment %in% "old"], "Prox1")

print("Number cells with >2 Prox1 in new experiments:")
cell_gene_count(sce[, sce$experiment %in% "new"], "Prox1")

print("Number cells with >2 Prox1 OR >2 Pdpn AND Pecam1 in old experiment:")
cell_gene_comb(sce[,sce$experiment %in% "old"], "Pdpn", "Pecam1", "Prox1")

print("Number cells with  >2 Prox1 OR >2 Pdpn AND Pecam1 in new experiment:")
cell_gene_comb(sce[,sce$experiment %in% "new"], "Pdpn", "Pecam1", "Prox1")


```

### Visualize new old experiment

#### Pdpn

```{r}
#| label: dimplots
p1 <- plotReducedDim(sce[,sce$experiment %in% "old"], 
                     dimred="TSNE", 
                     colour_by="Pdpn") + ggtitle("old")

p2 <- plotReducedDim(sce[,sce$experiment %in% "new"], 
                     dimred="TSNE", 
                     colour_by="Pdpn") + ggtitle("new")

wrap_plots(list(p1,p2), nrow = 1) + 
  plot_layout(guides = "collect") &
  theme(legend.position = "top")

```

##### Prox1

```{r}
#| label: dimplots prox1
p1 <- plotReducedDim(sce[,sce$experiment %in% "old"], 
                     dimred="TSNE", 
                     colour_by="Prox1") + ggtitle("old")

p2 <- plotReducedDim(sce[,sce$experiment %in% "new"], 
                     dimred="TSNE", 
                     colour_by="Prox1") + ggtitle("new")

wrap_plots(list(p1,p2), nrow = 1) + 
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

##### Pecam1

```{r}
#| label: dimplots pecam1
p1 <- plotReducedDim(sce[,sce$experiment %in% "old"], 
                     dimred="TSNE", 
                     colour_by="Pecam1") + ggtitle("old")

p2 <- plotReducedDim(sce[,sce$experiment %in% "new"], 
                     dimred="TSNE", 
                     colour_by="Pecam1") + ggtitle("new")

wrap_plots(list(p1,p2), nrow = 1) + 
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```
